---
title: "Extension Content: Divisive Clustering"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
---

```{r}
#| results: hide
#| message: false
#| warning: false
pkg_vec <- c("palmerpenguins", "cluster", "factoextra")
for (x in pkg_vec) {
  if (!requireNamespace(x, quietly = TRUE)) {
    install.packages(x)
  }
}
library(palmerpenguins)
library(cluster)
library(factoextra)
```

# Performance in various scenarios

## Well-separated clusters

```{r}
# Set seed for reproducibility
set.seed(123)

# 1. Well‑separated clusters
# ---------------------------
# Generate three well‑separated 2D Gaussian clusters
n <- 50
cl1 <- matrix(rnorm(2*n, mean = 0, sd = 0.3), ncol = 2)
cl2 <- matrix(rnorm(2*n, mean = 3, sd = 0.3), ncol = 2)
cl3 <- matrix(rnorm(2*n, mean = 6, sd = 0.3), ncol = 2)
data_well <- rbind(cl1, cl2, cl3)
```

```{r}
plot_tbl <- data_well |>
  tibble::as_tibble() |>
  dplyr::mutate(cluster = factor(rep(1:3, each = n)))
ggplot(plot_tbl, aes(x = V1, y = V2, col = cluster)) +
  geom_point() +
  labs(title = "Well-separated clusters") +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy")
```

```{r}
# Compute distance matrix
d_well <- dist(data_well)

# Plot dendrograms for average, complete, single linkage
par(mfrow = c(1,3), mar = c(4,4,2,1))
plot(hclust(d_well, method = "average"), main = "Well‑sep: Average", ylab = "Height")
plot(hclust(d_well, method = "complete"), main = "Well‑sep: Complete", ylab = "")
plot(hclust(d_well, method = "single"),   main = "Well‑sep: Single",   ylab = "")
```

## Poorly separated clusters

```{r}
# 2. Poorly separated clusters
# -----------------------------
# Generate three overlapping 2D Gaussian clusters
cl1_p <- matrix(rnorm(2*n, mean = 0, sd = 1), ncol = 2)
cl2_p <- matrix(rnorm(2*n, mean = 4, sd = 1.5), ncol = 2)
cl3_p <- matrix(rnorm(2*n, mean = 7, sd = 1.5), ncol = 2)
data_poor <- rbind(cl1_p, cl2_p, cl3_p)
```

```{r}
plot_tbl_poor <- data_poor |>
  tibble::as_tibble() |>
  dplyr::mutate(cluster = factor(rep(1:3, each = n)))
ggplot(plot_tbl_poor, aes(x = V1, y = V2, col = cluster)) +
  geom_point() +
  labs(title = "Poorly-separated clusters") +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy")
```

```{r}

d_poor <- dist(data_poor)

par(mfrow = c(1,3))
plot(hclust(d_poor, method = "average"), main = "Poor‑sep: Average", ylab = "Height")
plot(hclust(d_poor, method = "complete"), main = "Poor‑sep: Complete", ylab = "")
plot(hclust(d_poor, method = "single"),   main = "Poor‑sep: Single",   ylab = "")
```

## L-shaped cluster

```{r}
# 3. One “L”‑shaped cluster + one well‑sep cluster
# -------------------------------------------------
# Generate L‑shape: two arms along axes
arm1 <- cbind(seq(0, 5, length.out = n), rep(0, n))
arm2 <- cbind(rep(0, n), seq(0, 5, length.out = n))
L_cluster <- rbind(arm1, arm2) + matrix(rnorm(4*n, sd = 0.1), ncol = 2)

# Well‑sep cluster
cl_sep <- matrix(rnorm(2*n, mean = 2.5, sd = 0.3), ncol = 2)

data_L <- rbind(L_cluster, cl_sep)
```

```{r}
plot_tbl_L <- data_L |>
  tibble::as_tibble() |>
  dplyr::mutate(cluster = factor(c(rep(1, length(arm1) ), rep(2, n))))
ggplot(plot_tbl_L, aes(x = V1, y = V2, col = cluster)) +
  geom_point() +
  labs(title = "L-shaped cluster + well-separated cluster") +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy")
```

```{r}
d_L   <- dist(data_L)

par(mfrow = c(1,2))
plot(hclust(d_L, method = "average"), main = "L + sep: Average", ylab = "Height")
plot(hclust(d_L, method = "single"),  main = "L + sep: Single",  ylab = "")

# Reset plotting layout
par(mfrow = c(1,1))
```

# Divisive Clustering

Divisive clustering is a **top‑down** hierarchical approach in which you start with all observations in one cluster and recursively split until each cluster is sufficiently homogeneous. A simple algorithmic sketch is:

1. **Identify the “widest” cluster**  
   - Compute, for each current cluster, a measure of its spread (e.g. average pairwise dissimilarity or maximum dissimilarity).  
   - Select the cluster with the largest spread.

2. **Find the most “extreme” point**  
   - Within that cluster, choose the observation whose **average** dissimilarity to all other members is largest.

3. **Form a new subcluster**  
   - Remove that extreme point into its own cluster.  
   - Assign to it any point in the original cluster whose average dissimilarity to the extreme point is now **smaller** than its average dissimilarity to the remainder of its original cluster.

4. **Repeat**  
   - Recompute spreads for all clusters and go back to step 1 until no further splits occur (or until you reach a desired number of clusters).

> - Unlike agglomerative methods, divisive clustering allows you to focus splits where they matter most (the clusters with greatest internal heterogeneity).  
> - The “stop” criterion can be based on a minimum cluster size, a threshold for maximum within‑cluster dissimilarity, or simply running until a fixed number of clusters is obtained.

```{r}

# 1. Load and clean the penguins dataset
data("penguins", package = "palmerpenguins")
penguins_clean <- penguins |>
  dplyr::select(bill_length_mm,
                bill_depth_mm,
                flipper_length_mm,
                body_mass_g,
                species) |>
  na.omit()

penguins_mat <- as.matrix(penguins_clean[, -5])  # Exclude species column for clustering

# 2. Compute a (standardised) distance matrix
#    We standardise to put all measurements on comparable scales
dist_mat <- dist(scale(penguins_mat), method = "euclidean")

# 3. Perform divisive (DIANA) clustering
diana_res <- diana(dist_mat, diss = TRUE)

# 4. Plot the divisive dendrogram
plot(diana_res, 
     main = "DIANA Divisive Clustering on Palmer Penguins",
     ylab = "Height (dissimilarity)")

# 5. Cut the tree into k clusters (e.g. k = 3)
k <- 3
penguin_clusters <- cutree(as.hclust(diana_res), k = k)

# 6. Visualise the cluster partition in PC space
fviz_cluster(list(data = penguins_mat, cluster = penguin_clusters),
             geom = "point",
             ellipse.type = "convex",
             show.clust.cent = TRUE) +
  ggtitle(glue::glue("Penguins clustered by DIANA (k = {k})"))
```

```{r}
#| results: asis
# 7. Inspect cluster assignments
table(penguins_clean$species, penguin_clusters) |>
  knitr::kable()
```

# Ordering of samples

The dendrogram orders samples along the "x-axis" based on their hierarchical clustering structure.
It is worth briefly noting that one such method is to begin at the top and place the child node on the left that has the lower merge height.
I think this is what `R` uses in `plot(hist(...))` (`S` used it before and it looks like it), but I can't find it confirmed in help files.

```{r}
# Example of ordering samples in a dendrogram
hc <- hclust(dist(penguins_mat), method = "ward.D2")
plot(hc, main = "Hierarchical Clustering Dendrogram",
      xlab = "Samples (ordered by merge height)",
      ylab = "Height (dissimilarity)")
```

# CITRUS

# Aghaeepour (2013)