---
title: "Extension Content: Preprocessing data and additional proximity measures"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
---

```{r}
#| include: false
library(ggplot2)
library(dplyr)
library(purrr)
library(tidyr)
library(igraph)
library(FNN)
library(MASS)
```

# Standardising numeric data

Before calculating distance measures, it is often important to standardise numeric variables because:

- they may have different scales, e.g. one variable has a range of 0–100 while another has a range of 0–1000, or
- they may be measured in different units, e.g. height in cm and weight in kg, or

For example, suppose we have two variables measured in different units and so have very different scales:

| Observation | Income (Rs) | Weight (kg) |
| ----------- | ----------- | ----------- |
| A           | 170 000     | 65          |
| B           | 175 000     | 120         |
| C           | 160 000     | 60          |

```{r}
data_mat <- tibble::tibble(
  income = c(170000, 175000, 160000),
  weight = c(65, 120, 60)
) |>
  as.matrix()
rownames(data_mat) <- c("A", "B", "C")
```

Clearly, weight has little impact on the distance between observations:

```{r}
dist(data_mat, method = "euclidean")
```

So, we'll typically want to standardise these variables before calculating distances.

We'll discuss various options for standardising numeric variables, including:

- z-score standardisation
- range normalisation
- percentile transformation
- robust standardisation

## Option 1: Z-score Standardisation

For a numeric variable $X$ with observations $x_1, x_2, \dots, x_n$:

1. Compute the **sample mean**:

   $$
     \bar{x} = \frac{1}{n}\sum_{i=1}^n x_i.
   $$
2. Compute the **sample standard deviation**:

   $$
     s = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2}.
   $$
3. Transform each observation to its **z‑score**:

   $$
     z_i = \frac{x_i - \bar{x}}{s}.
   $$

After this transformation, each standardised variable has mean 0 and variance 1.
Distances based on these z‑scores are often referred to as **standardised Euclidean** distances.

```{r}
data_mat |>
  scale() 
```

```{r}
data_mat |>
  scale() |>
  dist(method = "euclidean") |>
  signif(3)
```

## Option 2: Range Normalisation

Range normalisation rescales each variable to a \[0,1\] range, which is useful when variables have different scales or units.

For observation $i$ in variable $X$ with values $x_1,\dots,x_n$:

1. Compute the **minimum** and **maximum**:

   $$
     \min(X) = \min(x_1, \dots, x_n), \quad \max(X) = \max(x_1, \dots, x_n).
   $$

2. Transform each observation to its **normalised value**:
    $$
      x_i' = \frac{x_i - \min(X)}{\max(X) - \min(X)}.
    $$

This transformation ensures:

- the minimum value becomes 0 and the maximum becomes 1.
  - This can be important when some variables are non-numeric, as the variables all have the same range, making them comparable.
- the impact of extreme outliers is reduced, as they are compressed into the \[0,1\] range.
  - However, extreme outliers can still distort distances if they "squash" all other observations for that variable close together.
- This is the approach taken by Gower's dissimilarity measure for numeric variables.

```{r}
data_mat_scale <- data_mat |>
  apply(2, function(x) (x - min(x)) / (max(x) - min(x)))
data_mat_scale |>
  signif(2)
```

```{r}
data_mat_scale |>
  dist(method = "euclidean") |>
  signif(3)
```

## Option 3: Percentile Transformation

Converting values to percentile ranks redistributes data onto a uniform \[0,1] scale based on their empirical cumulative distribution function.
For observation $i$ in variable $X$ with values $x_1,\dots,x_n$:

1. Compute the **rank** of $x_i$ (e.g.\ average rank for ties):

   $$
     r_i = \#\{j : x_j \le x_i\}.
   $$
2. Convert to **percentile**:

   $$
     p_i = \frac{r_i - 1}{n - 1},
   $$

   so that the minimum becomes 0 and the maximum 1.

This approach:

* **Reduces impact of extreme outliers** by compressing the tails.
* **Ensures a uniform distribution** on \[0,1], making subsequent distances less influenced by skew.

However, the loss of shape information means it may not capture relative distances as well as z-scores or even range normalisation.

This is the same as the rank approach for ordinal variables, but applied to numeric data.

```{r}
data_mat_scaled <- apply(data_mat, 2, function(x) {
  rank(x, ties.method = "average") / length(x)
})
data_mat_scaled |>
  signif(2)
```

```{r}
data_mat_scaled |>
  dist(method = "euclidean") |>
  signif(3)
```

### Option 4: Log transformation

Sometimes a simple monotonic transform can stabilise variance and reduce skew before standardisation.

For strictly positive $x_i$, the log transform is:

$$
  y_i = \log(x_i)
$$

* **Effect:** compresses large values, spreads small values.
* **When to use:** data with multiplicative effects or heavy right skew (e.g. incomes, biological measures).

```{r}
x <- c(1, 10, 100, 1000)
y_log <- log10(x)
y_log
```

#### Extra options

##### Box–Cox transformation

The Box–Cox family generalises logs with a parameter $\lambda$:

$$
  y_i^{(\lambda)} = \begin{cases}
    \frac{x_i^{\lambda} - 1}{\lambda}, & \lambda \neq 0, \\
    \log(x_i), & \lambda = 0.
  \end{cases}
$$

* **Effect:** for $\lambda>0$, similar to power transforms; as $\lambda\to0$, approaches log.
* **Choosing $\lambda$:** often by maximum-likelihood or cross-validation to best approximate normality.

```{r}
# Estimate lambda for x
bc <- boxcox(lm(x ~ 1), plotit = FALSE)
lambda_hat <- bc$x[which.max(bc$y)]
lambda_hat
```

##### `asinh` transformation

One can use the inverse `asinh` transformation for data that includes zero or negative values:

```{r}
x <- c(-100, -10, 0, 1, 10, 100)
y_asinh <- asinh(x)
y_asinh
```

# Distance measures

## Choice of $p$ for Minkowski distance

The Minkowski distance is defined as:

$$
  d_p(x, y) = \left( \sum_{i=1}^n |x_i - y_i|^p \right)^{1/p}
$$

Special cases include:

- $p=1$: **Manhattan distance** (or city block distance)
- $p=2$: **Euclidean distance**
- $p=\infty$: **Chebyshev distance** (maximum absolute difference)

Plotted on a 2D plane, these distances form different shapes:

```{r}


# 1. Define norms to plot
ps <- c(1, 2, Inf)

# 2. Build a dense grid of (x,y) in [-1,1]^2
grid <- expand.grid(
  x = seq(-1, 1, length.out = 400),
  y = seq(-1, 1, length.out = 400)
) %>% as_tibble()

# 3. For each p, compute d_p and keep points with |d_p - 1| small
plot_tbl <- map_df(ps, function(pval) {
  grid %>%
    mutate(
      dist = if (is.infinite(pval)) {
        pmax(abs(x), abs(y))
      } else {
        (abs(x)^pval + abs(y)^pval)^(1/pval)
      }
    ) %>%
    filter(abs(dist - 1) < 0.003) %>%      # tolerance band
    mutate(
      norm = factor(
        paste0("p = ", ifelse(is.infinite(pval), "∞", pval)),
        levels = c("p = 1", "p = 2", "p = ∞")
      )
    )
})

# 4. Plot
p <- ggplot(plot_tbl, aes(x, y)) +
  geom_point(size = 0.4, alpha = 0.7) +
  facet_wrap(~ norm, nrow = 1) +
  coord_equal() +
  labs(
    x = expression(x[1]),
    y = expression(x[2])
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )
path_p  <- "_tmp/fig/unit_circle_norms.png"
if (!dir.exists("_tmp/fig")) dir.create("_tmp/fig", recursive = TRUE)
ggsave(path_p, p, width = 8, height = 3.5)
```

```{r}
#| results: asis
#| echo: false
knitr::include_graphics(path_p)
```

## Comparing sequences

When observations are **sequences** (e.g. strings), one common metric is:

* **Edit (Levenshtein) distance:** the minimum number of insertions, deletions or substitutions needed to transform one string into another.

```{r}
library(stringdist)
# Compute Levenshtein distance
stringdist("kitten", "sitting", method = "lv")
```

## Graph‑Based Distances

In high dimensions or on complex manifolds, direct distances may lose contrast.
Graph‑based approaches restore local structure:

1. **Construct a k‑nearest‑neighbour (kNN) graph**:

   * Each point is a node.
   * Connect node $i$ to its $k$ closest points by Euclidean distance.

2. **Compute shortest‑path distances** on this graph:

   * Use an algorithm to find minimal cumulative edge weight between nodes.

### 5.1 R Example with `igraph` and `FNN`

```{r}
# Prepare data
set.seed(123)
data(iris)
X <- as.matrix(iris[, 1:4])
n <- nrow(X)
k <- 5

# 1. Find kNN indices (matrix n x k)
knn_info <- get.knn(X, k = k)

# 2. Build edge list robustly
edges <- map_dfr(
  seq_len(n),
  function(i) {
    nbrs <- knn_info$nn.index[i, ]
    # compute Euclidean distances
    dists <- sqrt(rowSums((X[nbrs, , drop=FALSE] - X[i, ])^2))
    tibble(
      from = i,
      to = nbrs,
      weight = dists
    )
  }
)

# 3. Create undirected graph
g <- graph_from_data_frame(edges, directed = FALSE)

# 4. Compute graph-based distances
dist_matrix <- distances(g, weights = E(g)$weight)

# Inspect a subset
dist_matrix[1:5, 1:5]
```

### Plot the kNN Graph

```{r}
# Restrict to first 50 nodes for clarity
subg <- induced_subgraph(g, 1:50)
# Layout in 2D via PCA for plotting
coords <- prcomp(X[1:50, ])$x[, 1:2]
V(subg)$x <- coords[,1]
V(subg)$y <- coords[,2]

plot(
  subg,
  layout = as.matrix(coords),
  vertex.size = 5,
  vertex.label = NA,
  edge.color = "grey",
  main = "kNN Graph (Iris, first 50 observations)"
)
```

This visualisation shows each point connected to its five nearest neighbours in the original feature space, highlighting local neighbourhood structure that informs graph‑based distances.

# High-dimenensional data

In high dimensions ($p$ large):

* Distances tend to **concentrate** (most pairs end up similarly far).
* **Noise and irrelevant features** can dominate meaningful signal.

**Alternative preprocessing:**

* **Principal Component Analysis (PCA):** project onto top $k$ components capturing most variance, reducing $p$ to $k\ll p$, speeding up computations and mitigating noise.
* **Feature selection** to remove uninformative variables.



