---
title: Clustering
subtitle: Part One
date: today
format:
  pdf:
    embed-resources: true
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
---

# Introduction

The `data_tidy_cell_lineage` dataset contains measurements of various cell markers collected from cytometry experiments. These markers are used to classify the cells into different lineages. In this assignment, you will explore clustering techniques to identify sub-populations of cells and compare different clustering methods.

The dataset can be loaded as follows:

```{r}
#| results: hide
#| warning: false
#| message: false
pkg_vec_cran <- c("remotes", "tibble", "ggplot2", "cowplot")
for (x in pkg_vec_cran) {
  if (!requireNamespace(x, quietly = TRUE)) {
      install.packages(x)
  }
}
library(tibble)
remotes::install_github("MiguelRodo/DataTidy24STA5069Z@2025")
data("data_tidy_cell_lineage", package = "DataTidy24STA5069Z")
data_tidy_cell_lineage
```

@fig-cell-lineage is an exploratory figure that displays density plots of the cell markers.

```{r}
#| include: false
path_cytof_fig <- projr::projr_path_get("raw-data-img", "p-density-all-lineage.png")
```

::: {#fig-cell-lineage}
![](`r path_cytof_fig`)

*Density plots of the markers across the cells.*
:::

# Questions

## Question One: Dissimilarity measures (6 marks)

1. **Write** an R function `gower_dist(x)` that accepts a `data.frame` (numeric + factor columns + ordinal columns) and returns the Gower dissimilarity matrix. (6 marks) 
2. **Apply** your `gower_dist()` to the `car` dataset below, and also compute `cluster::daisy(car, metric = "gower")`.  
3. **Verify** that your `gower_dist()` function returns the same result as `cluster::daisy(car, metric = "gower")`. (2 marks)

```r
car <- data.frame(
  horsepower    = c(130, 250, 180, 95, 220, 160, 300, 140, 200, 175),
  weight        = c(1200, 1650, 1420, 980, 1580, 1350, 1720, 1250, 1500, 1380),
  price_bracket = factor(c(
    "low", "high", "mid", "low", "high", 
    "mid", "high", "low", "mid", "mid"
  )),
  colour        = factor(c(
    "red", "black", "blue", "white", "silver", 
    "blue", "black", "red", "white", "silver"
  ))
)
```

## Question Two: Hierarchical Clustering (11 marks)

a. **R code**: Perform hierarchical clustering on the `data_tidy_cell_lineage` dataset using a distance measure and linkage approach you think is likely appropriate. (1 mark)
b. **R code + plot + typed response**: Motivate the optimal number of clusters based on:
   - **i. the dendrogram** (2 marks)
   - **ii. the silhouette statistic** (2 marks)
c. **R code**: Cut the dendrogram to obtain cluster assignments. (1 mark)
d. **R code + plot**: Create a scatter plot of the first two MDS dimensions of the cell lineage data, coloring the points by their cluster assignments. Comment on the results. Below is code to create the MDS assignments. (3 marks)

```{r}
#| eval: false
if (!requireNamespace("smacof", quietly = TRUE)) {
  install.packages("smacof")
}
set.seed(123)
dist_mat_max <- dist(
  data_tidy_cell_lineage |>
    dplyr::select(-cell_id),
  method = "maximum")
smacof_mds_obj <- smacof::mds(
  delta = dist_mat_max, # dissimilarities
  ndim = 2, # desired dimension of configuration
  type = "ordinal", # type of dissimilarity
  init = "torgerson" # initialise with classical scaling
)
plot_tbl <- data_tidy_cell_lineage |>
  dplyr::mutate(
    v1 = smacof_mds_obj[["conf"]][, 1], 
    v2 = smacof_mds_obj[["conf"]][, 2],
    cluster = clust_vec
  )
```

e. **Typed response**: Typically, downstream analyses assume that cluster assignments are known with certainty. Comment on the validity of this assumption in the context of the cell lineage data. (2 marks)
f. **Plot**: Perform hierarchical clustering on the `data_tidy_cell_lineage` dataset using the following approaches, creating the same plot as in part d for each approach (no need to consider the number of clusters in depth):
   - **i. complete linkage, Chebychev distance**: (4 marks)
   - **ii. single linkage, Euclidean distance**: (4 marks)
g. **Typed response**: Compare the results of the clustering approaches in parts d and f. Which approach do you think is most appropriate for this dataset? What aspects of the algorithms do you think led to that? (2 marks)