---
title: Clustering
subtitle: Part One
date: today
format:
  html:
    embed-resources: true
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    toc: true
    toc-depth: 2
---

# Introduction

Use the `data_tidy_cell_lineage` dataset for this question, which can be loaded as follows:


```{r}
#| results: hide
#| warning: false
#| message: false
pkg_vec_cran <- c("remotes", "tibble", "ggplot2", "cowplot")
for (x in pkg_vec_cran) {
  if (!requireNamespace(x, quietly = TRUE)) {
      install.packages(x)
  }
}
library(tibble)
remotes::install_github("MiguelRodo/DataTidy24STA5069Z@2025")
data("data_tidy_cell_lineage", package = "DataTidy24STA5069Z")
data_tidy_cell_lineage
```

```{r}
library(ggplot2)
path_dir_fig <- file.path("_tmp", "fig", "prac1")
if (dir.exists(path_dir_fig)) {
  unlink(path_dir_fig, recursive = TRUE)
}
dir.create(path_dir_fig, recursive = TRUE)
```

Note that the package needs to be updated to the latest version to load the data in this manner (via the `data` command rather than reading in a CSV file).

@fig-cell-lineage is an exploratory figure that displays density plots of the cell markers.

```{r}
#| include: false
path_cytof_fig <- projr::projr_path_get("raw-data-img", "p-density-all-lineage.png")
```

::: {#fig-cell-lineage}
![](`r path_cytof_fig`)

*Density plots of the markers across the cells.*
:::

## Question One: Dissimilarity measures (6 marks)

1. **Write** an R function `gower_dist(x)` that accepts a `data.frame` (numeric + factor columns + ordinal columns) and returns the Gower dissimilarity matrix. (6 marks) 

```{r}
# 1. Define the Gower distance function
gower_dist <- function(x) {
  x <- as.data.frame(x)
  n <- nrow(x)
  p <- ncol(x)
  # initialise total dissimilarity matrix
  D <- matrix(0, n, n)
  
  for (j in seq_len(p)) {
    col <- x[[j]]
    
    # Numeric columns
    if (is.numeric(col)) {
      rng <- max(col, na.rm = TRUE) - min(col, na.rm = TRUE)
      # if every element is the same, then the range is 0,
      # and so it contributes nothing to the dissimilarity
      if (rng == 0) {
        dmat <- matrix(0, n, n)
      } else {
        dmat <- abs(outer(col, col, "-")) / rng
        # can also loop, as follows:
        # dmat <- matrix(0, n, n)
        # for (i in seq_len(n)) {
        #   for (k in seq_len(n)) {
        #     dmat[i, k] <- abs(col[i] - col[k]) / rng
        #   }
        # }
        # the `outer` function takes less typing.
        # what it does is that it creates
        # an m x n matrix (where the first argument is a vector of length m and the second is a vector of length n)
        # and fills it with the results of the function ("-", in this case)
        # applied to each pair of elements.
        # The ij-th element is the result of applying the function to the
        # i-th element of the first vector and the j-th element of the second vector.
        # In this case, we subtract the j-th element of the second vector from the i-th element of the first vector.
        # The `abs` and `/ rng` parts are applied after the subtraction,
        # and ensure that the result is positive and normalised by the range.
      }
      
    # Ordinal factors: treat as numeric codes normalised by range
    } else if (is.ordered(col)) {
      codes <- as.numeric(col)

      if (rng == 0) {
        dmat <- matrix(0, n, n)
      } else {
        # calculate z_ij
        z_vec <- (codes - min(codes)) / max(codes - min(codes))
        dmat <- abs(outer(z_vec, z_vec, "-")) / (max(z_vec) - min(z_vec))
      }
      
    # Nominal factors: 0 if same level, 1 if different
    } else if (is.factor(col)) {
      dmat <- outer(col, col, FUN = Vectorize(function(a, b) as.numeric(a != b)))
      
    } else {
      stop("Unsupported column type: ", class(col))
    }
    
    D <- D + dmat
  }
  
  # Average over all variables
  D <- D / p
  rownames(D) <- rownames(x)
  colnames(D) <- rownames(x)
  D
}
```

2. **Apply** your `gower_dist()` to the `car` dataset below, and also compute `cluster::daisy(car, metric = "gower")`.  

```{r}
car <- data.frame(
  horsepower    = c(130, 250, 180, 95, 220, 160, 300, 140, 200, 175),
  weight        = c(1200, 1650, 1420, 980, 1580, 1350, 1720, 1250, 1500, 1380),
  price_bracket = factor(c(
    "low", "high", "mid", "low", "high", 
    "mid", "high", "low", "mid", "mid"
  )),
  colour        = factor(c(
    "red", "black", "blue", "white", "silver", 
    "blue", "black", "red", "white", "silver"
  ))
)
gower_dist(car)
```

3. **Verify** that your `gower_dist()` function returns the same result as `cluster::daisy(car, metric = "gower")`. (2 marks)

```{r}
D_my    <- gower_dist(car)
D_daisy <- as.matrix(cluster::daisy(car, metric = "gower"))
# use all.equal instead of `identical` to allow for small numerical differences,
# and also in case there's a difference in the class of the objects.
identical <- all.equal(D_my, D_daisy)
identical
```

Note that we used `all.equal` instead of `identical` to allow for small numerical differences.
We can see that `identical` is `FALSE` below:
```{r}
identical(D_my, D_daisy)
```

They seem to have precisely the same structure:
```{r}
str(D_my)
str(D_daisy)
```

Therefore, the differences are likely due to numerical precision.
We can check for differences in each element:

```{r}
for (i in seq_len(nrow(D_my))) {
  for (j in seq_len(ncol(D_my))) {
    if (D_my[i, j] != D_daisy[i, j]) {
      message("Mismatch at row ", i, ", column ", j, ": ", D_my[i, j], " vs ", D_daisy[i, j])
    }
  }
}
```

If we compare only at the fourth decimal place, we can see that they are indeed the same:

```{r}
identical(
  signif(D_my, digits = 4),
  signif(D_daisy, digits = 4)
)
```

One could have simply also summed over the differences:

```{r}
sum(abs(D_my - D_daisy))
```

They're minimal, and so we can conclude that the two matrices are essentially the same.
This would have worked even if the matrices had different attributes, such as row and column names, as long as the numerical values are the same.

# Question two: hierarchical clustering (11 marks)

a. **R code**: Perform hierarchical clustering on the `data_tidy_cell_lineage` dataset. (1)

```{r}
dist_mat <- dist(
    data_tidy_cell_lineage |>
      dplyr::select(-cell_id),
    method = "euclidean"
  )
hclust_obj <- hclust(dist_mat, method = "ward.D2")
```

b. **R code + plot + typed response**: Motivate the optimal number of clusters based on the:
- i. **R code + plot + typed response**: dendrogram. (2)

```{r}
#| results: hide
path_dendro <- file.path(
  "_tmp", "fig", "prac1", "p-q2-b-i.png"
)
png(path_dendro, width = 10, height = 10, units = "cm", res = 300)
plot(hclust_obj)
dev.off()
```

```{r, results='asis'}
knitr::include_graphics(path_dendro)
```

- ii. **R code + output + typed response**: silhouette statistic. (2)

```{r}
sil_stat_3 <- cluster::silhouette(
  cutree(hclust_obj, k = 3),
  dist(
    data_tidy_cell_lineage |>
      dplyr::select(-cell_id),
    method = "euclidean"
  )
)
sil_stat_4 <- cluster::silhouette(
  cutree(hclust_obj, k = 4),
  dist(
    data_tidy_cell_lineage |>
      dplyr::select(-cell_id),
    method = "euclidean"
  )
)
summary(sil_stat_3)
summary(sil_stat_4)
```

c. **R code**: Cut the dendrogram to obtain cluster assignments. (1)

```{r}
clust_vec <- cutree(hclust_obj, k = 3)
```

d. **R code + plot**: Create a scatter plot of the first two MDS dimensions of the cell lineage data, coloring the points by their cluster assignments. Comment on the results. Below is code to create the MDS assignments. (3)

```{r}
set.seed(123)
dist_mat_max <- dist(
  data_tidy_cell_lineage |>
    dplyr::select(-cell_id),
  method = "maximum")
smacof_mds_obj <- smacof::mds(
  delta = dist_mat_max, # dissimilarities
  ndim = 2, # desired dimension of configuration
  type = "ordinal", # type of dissimilarity
  init = "torgerson" # initialise with classical scaling
)
```

```{r}
plot_tbl <- data_tidy_cell_lineage |>
  dplyr::mutate(
    v1 = smacof_mds_obj[["conf"]][, 1], 
    v2 = smacof_mds_obj[["conf"]][, 2],
    cluster = as.character(clust_vec)
  )
p_mds <- ggplot(
  plot_tbl,
  aes(x = v1, y = v2, col = cluster)
) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  geom_point() +
  labs(
    x = "MDS1",
    y = "MDS2"
  ) +
  theme(legend.title = element_blank())
path_plot <- file.path(
  path_dir_fig, "p-q1-d.png"
)
cowplot::ggsave2(
  path_plot,
  p_mds,
  width = 10,
  height = 10,
  units = "cm"
)
```

```{r, results='asis'}
knitr::include_graphics(path_plot)
```

e. **Typed response**: Typically, downstream analyses assume that cluster assignments are known with certainty. Comment on the validity of this assumption in the context of the cell lineage data. (2)

It is not clear from the above analysis that there are not more clusters present and some cells may be misclassified ($\checkmark$), in particular cells from the bright yellow cluster.
Therefore this assumption may typically be false ($\checkmark$), but may be a useful simplification for some analyses.

f. **Plot**: Perform hierarchical clustering on the `data_tidy_cell_lineage` dataset using the following approaches, creating the same plot as in part d for each approach (no need to consider the number of clusters in depth):
   - **i. complete linkage, Chebychev distance**: (4 marks)
   - **ii. single linkage, Euclidean distance**: (4 marks)

For part i:

```{r}
dist_mat <- dist(
    data_tidy_cell_lineage |>
      dplyr::select(-cell_id),
    method = "maximum"
  )
hclust_obj <- hclust(dist_mat, method = "complete")
```


```{r}
#| results: hide
path_dendro <- file.path(
  "_tmp", "fig", "prac1", "p-q2-f-i-dendro.png"
)
png(path_dendro, width = 10, height = 10, units = "cm", res = 300)
plot(hclust_obj)
dev.off()
```

Not an especially clear number of clusters, but we can choose 3 to be consistent with the previous analysis.

```{r}
clust_vec <- cutree(hclust_obj, k = 3)
plot_tbl <- plot_tbl |>
  dplyr::mutate(
    cluster = clust_vec |> as.character()
  )
p_mds <- ggplot(
  plot_tbl,
  aes(x = v1, y = v2, col = cluster)
) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  geom_point() +
  labs(
    x = "MDS1",
    y = "MDS2"
  ) +
  theme(legend.title = element_blank())
path_plot <- file.path(
  path_dir_fig, "p-q2-f-i-scatter.png"
)
cowplot::ggsave2(
  path_plot,
  p_mds,
  width = 10,
  height = 10,
  units = "cm"
)
```

```{r, results='asis'}
knitr::include_graphics(path_plot)
```

The results are, interestingly, quite similar to the previous analysis, despite depending on only one marker for each difference.
Presumably this is because the markers driving the variation in the first plot are also the markers that have the greatest range, meaning they will still be the markers picked up by the Chebychev distance.

For part ii:
```{r}
dist_mat <- dist(
    data_tidy_cell_lineage |>
      dplyr::select(-cell_id),
    method = "euclidean"
  )
hclust_obj <- hclust(dist_mat, method = "single")
``` 

```{r}
#| results: hide
#| warning: false
path_dendro <- file.path(
  "_tmp", "fig", "prac1", "p-q2-f-ii-dendro.png"
)
png(path_dendro, width = 10, height = 10, units = "cm", res = 300)
plot(hclust_obj)
dev.off()
```

We could select three clusters, as before, but it is absolutely unclear that this is the right number of clusters and almost all the observations would be in the same cluster.
So, we will select ten clusters, and see what happens.

```{r}
plot_tbl <- plot_tbl |>
  dplyr::mutate(
    cluster = cutree(hclust_obj, k = 10) |>
      as.character()
  )
p_mds <- ggplot(
  plot_tbl,
  aes(x = v1, y = v2, col = cluster)
) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  ) +
  geom_point() +
  labs(
    x = "MDS1",
    y = "MDS2"
  ) +
  theme(legend.title = element_blank())
path_plot <- file.path(
  path_dir_fig, "p-q2-f-ii-scatter.png"
)
cowplot::ggsave2(
  path_plot,
  p_mds,
  width = 10,
  height = 10,
  units = "cm"
)
```

```{r}
knitr::include_graphics(path_plot)
```

g. **Typed response**: Comment on the differences between the clustering approaches and the results obtained. (2)
The first (Ward.D2 linkage + Euc distance) and second (complete linkage + Chebychev) approaches yield similar results that match the MDS representation closely.
Using single linkage in the third approach, however, discovers no clustering structure, as all points are essentially in the same cluster.

The first approach is likely more appropriate for this dataset, as it captures the variation in the markers more effectively than the second approach, which uses a single marker for each difference.
If the markers driving the variation captured by MDS are not the markers with the greatest range and variation, or there are several such markers with similar ranges, then the Chebychev distance may not capture the variation effectively.
The single linkage approach performs worst because the clusters are not sufficiently separated to avoid chaining, which is a common issue with single linkage clustering.