---
title: "Prac: Implementing Prototype Clustering Methods"
date: "2025-02-20"
format:
  html:
    self-contained: true
toc: true
toc-depth: 2
---

## Overview

In this practical, you will use the DBSCAN algorithm (from the `dbscan` R package) to investigate how the *eps* (the maximum neighborhood radius) affect the clustering outcome. You will:

- Generate or load a dataset.
- Create a k‑nearest neighbor (k-NN) distance plot to help choose an appropriate *eps*.
- Experiment with different *minPts* values.
- Visualize the clustering results and discuss the impact of your parameter choices.
- Discuss the strengths and weaknesses of DBSCAN in relation to the datasets you explore.

## Instructions

1. **Data Preparation:**

Load the following seven datasets, each simulating different clustering scenarios:

```{r}
#| eval: false
# Simulation of seven datasets for DBSCAN parameter exploration

set.seed(123)
library(MASS)
library(tibble)
library(ggplot2)
library(dbscan)

n <- 200

# 1. High‐separation spherical clusters
ds_high_sep <- rbind(
  mvrnorm(n/2, mu = c(0, 0), Sigma = diag(2)),
  mvrnorm(n/2, mu = c(5, 5), Sigma = diag(2))
)

# 2. High‐separation uniform clusters
ds_high_sep_uniform <- rbind(
  cbind(runif(n/2, 0, 1), runif(n/2, 0, 1)),
  cbind(runif(n/2, 4, 5), runif(n/2, 4, 5))
)

# 3. Low‐separation spherical clusters
ds_low_sep <- rbind(
  mvrnorm(n/2, mu = c(0, 0), Sigma = diag(2)),
  mvrnorm(n/2, mu = c(2, 2), Sigma = diag(2))
)

# 4. Weirdly shaped clusters (two half-moons)
n_moon <- 200
theta1 <- runif(n_moon/2, 0, pi)
moon1  <- cbind(x = cos(theta1),      y = sin(theta1))
theta2 <- runif(n_moon/2, 0, pi)
moon2  <- cbind(x = 1 - cos(theta2),  y = -sin(theta2) + 0.5)
ds_weird <- rbind(moon1, moon2)

# 5. High‐separation clusters + 10% uniform noise
noise_prop <- 0.25
noise_n    <- round(n * noise_prop)
x_range    <- apply(ds_high_sep, 2, range)
noise <- cbind(
  x = runif(noise_n, x_range[1,1], x_range[2,1]),
  y = runif(noise_n, x_range[1,2], x_range[2,2])
)
ds_high_sep_noise <- rbind(ds_high_sep, noise)

# 6. Medium-separation clusters with spherical distribution
ds_med_sep_spherical <- rbind(
  mvrnorm(n/2, mu = c(0, 0), Sigma = diag(2)),
  mvrnorm(n/2, mu = c(3, 3), Sigma = diag(2))
)

# 7. Medium-separation clusters with noise
noise_prop <- 0.25
noise_n    <- round(n * noise_prop)
x_range    <- apply(ds_med_sep_spherical, 2, range)
noise_med <- cbind(
  x = runif(noise_n, x_range[1,1], x_range[2,1]),
  y = runif(noise_n, x_range[1,2], x_range[2,2])
)
ds_med_sep_noise <- rbind(ds_med_sep_spherical, noise_med)

# Convert each to a tibble with columns x and y
high_sep        <- as_tibble(ds_high_sep,       .name_repair = ~c("x","y"))
low_sep         <- as_tibble(ds_low_sep,        .name_repair = ~c("x","y"))
weird_shapes    <- as_tibble(ds_weird,          .name_repair = ~c("x","y"))
high_sep_noise  <- as_tibble(ds_high_sep_noise, .name_repair = ~c("x","y"))
high_sep_uniform <- as_tibble(ds_high_sep_uniform, .name_repair = ~c("x","y"))
med_sep_spherical <- as_tibble(ds_med_sep_spherical, .name_repair = ~c("x","y"))
med_sep_noise <- as_tibble(ds_med_sep_noise, .name_repair = ~c("x","y"))
```

2. **Applying DBSCAN:**

- Use the `kNNdistplot()` function from the `dbscan` package to visualize the k-NN distances.
- Choose appropriate values for *eps* based on the k-NN distance plot.
- Apply the `dbscan()` function to perform clustering.
- Tweak the parameters *eps* and *minPts* to see how they affect the clustering results.
- Display the results.

The dimensionality is the same across datasets, so we'll set `minPts` to 5 throughout the exercises.

```{r}
#| eval: false
#| include: false
# here we save all the plots in a temporary directory, 
# _tmp/prac3/dbscan.
# each dataset has two plots saved for it, namely
# p-<data_name>-kNNdistplot.png and p-<data_name>-dbscan.png
library(dbscan)
# clean any previous temporary directories
path_tune_dir <- file.path("_tmp", "prac3", "tune")
if (dir.exists(path_tune_dir)) {
  unlink(path_tune_dir, recursive = TRUE)
}
dir.create(path_tune_dir, recursive = TRUE)

plot_knn_and_dbscan <- function(.data,
                                data_name,
                                minPts, 
                                eps = NULL,
                                path_dir_save) {
  n <- nrow(.data)
  
  # k-NN distance plot
  path_p <- file.path(path_dir_save, paste0("p-", data_name, "-kNNdistplot.png"))
  png(path_p, width = 800, height = 600)
  kNNdistplot(.data, k = 2 * ncol(.data))
  abline(h = eps, col = "red", lty = 2)  # example threshold for eps
  dev.off()
  
  # Run DBSCAN with chosen parameters
  if (is.null(eps)) {
    message("Please provide a value for eps.")
    return(NULL)
  }
  db_res <- dbscan(.data, eps = eps, minPts = minPts)
  
  # Prepare data for plotting
  df <- as_tibble(.data)
  colnames(df) <- c("x", "y")
  df$cluster <- factor(db_res$cluster)
  
  # Plot DBSCAN results
  p <- ggplot(df, aes(x = x, y = y, color = cluster)) +
    geom_point(size = 2) +
    scale_colour_brewer(palette = "Set1") +
    ggtitle(paste("DBSCAN Clustering Results for", data_name))
  
  path_p <- file.path(path_dir_save, paste0("p-", data_name, "-dbscan.png"))
  ggplot2::ggsave(path_p, plot = p, width = 8, height = 6)
}

plot_knn_and_dbscan(
  high_sep, 
  data_name = "high_sep", 
  minPts = 5, 
  eps = 0.8, 
  path_dir_save = path_tune_dir
)

# it's very difficult to separate the clusters, with
# the actual clusters splitting when eps is too small
plot_knn_and_dbscan(
  low_sep, 
  data_name = "low_sep", 
  minPts = 5, 
  eps = 0.45, 
  path_dir_save = path_tune_dir
)

# less straightforward than one might have thought
# to separate the two half-moon shapes
plot_knn_and_dbscan(
  weird_shapes, 
  data_name = "weird_shapes", 
  minPts = 5, 
  eps = 0.3, 
  path_dir_save = path_tune_dir
)

# some noise points get classified as clusters,
# but the main clusters are still identified
plot_knn_and_dbscan(
  high_sep_noise, 
  data_name = "high_sep_noise", 
  minPts = 5, 
  eps = 1.1,
  path_dir_save = path_tune_dir
)

# medium-separation clusters with spherical distribution
# can work, but quite sensitive to eps
plot_knn_and_dbscan(
  med_sep_spherical,
  data_name = "med_sep_spherical",
  minPts = 5,
  eps = 0.6,
  path_dir_save = path_tune_dir
)

# medium-separation clusters with noise
# surprisingly robust
plot_knn_and_dbscan(
  med_sep_noise,
  data_name = "med_sep_noise",
  minPts = 5,
  eps = 0.6,
  path_dir_save = path_tune_dir
)
```

4. **Comparison with single-linkage hierarchical clustering:**  

Compare the results of DBSCAN with those from single-linkage hierarchical clustering using the `hclust()` function on the `med_sep_noise` dataset.

```{r}
#| eval: false
#| include: false
path_dir_hclust <- file.path("_tmp", "prac3", "hclust")
if (dir.exists(path_dir_hclust)) {
  unlink(path_dir_hclust, recursive = TRUE)
}
dir.create(path_dir_hclust, recursive = TRUE)

hclust_obj <- hclust(dist(med_sep_noise), method = "single")
path_p_dendro <- file.path(path_dir_hclust, "dendrogram-med_sep_noise.png")
png(path_p_dendro, width = 800, height = 600)
plot(hclust_obj, labels = FALSE, main = "Single-Linkage Hierarchical Clustering")
dev.off()

# Prepare data for plotting
df <- as_tibble(med_sep_noise)
colnames(df) <- c("x", "y")
df$cluster <- cutree(hclust_obj, k = 5) |>
  as.character() # assuming 5 clusters

# Plot DBSCAN results
p <- ggplot(df, aes(x = x, y = y, color = cluster)) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  ggtitle(paste("HClust Clustering Results for medium-separation noise"))

path_p <- file.path(path_dir_hclust, paste0("p-", "med_sep_noise", "-hclust.png"))
ggplot2::ggsave(path_p, plot = p, width = 8, height = 6)
```

5. **Discussion:**  
  
Write a brief summary on how parameter choices influence DBSCAN’s performance.

   ```{r}
   #| include: false
   # - Can be hard to divine what the right eps should be.
   # - Quite robust to noise, unlike hclust with single linkage.
   ```

6. **Scenario comparison:**

Compare the clustering results across the different datasets you created. Discuss how the shape and separation of clusters affect the choice of parameters, and relate that to the DBSCAN algorithm's strengths and weaknesses.

7. **Propose alternative interesting datasets:**  

Suggest other datasets or scenarios where DBSCAN might be particularly effective or ineffective, motivating your choices based on the characteristics of the data.
